---
title: "First_Year_Project"
output: pdf_document
date: "2023-02-07"
---

```{r setup, include=FALSE, results=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggpmisc")) install.packages("ggpmisc")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("ggsignif")) install.packages("ggsignif")
library("ggplot2")
library("ggpmisc")
library("ggpubr")
library("ggsignif")
library(tibble)
library(ggplot2)
library(broom)
```

# Question one
*Begin with describing and fitting a full model in which the intercepts and slopes of the extinction times versus numbers of pairs may be different in all four combinations of size and migratory status.*  
We start by importing the dataset directly from the csv file and saving it to the data variable:

```{r}
head(data <- read.csv('Factors Affecting Extinction.csv', header=T))
```

There are four different combinations of Size and Status, LR, LM, SR and SM. If we want to find the correlation between *extinction time* as a function of *pairs*, we can make a regression line with *pairs* as the predictor value and *extinction time* as the predicted value.

```{r}
ggplot(data, aes(x = Pairs, y = Time)) +
  geom_point() +
  facet_grid(Size ~ Status) +
  theme(legend.position = "top") +
  geom_smooth(method = "lm", formula = y ~ x) +
  
  stat_poly_eq(formula = y ~ x,
  aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
  parse = TRUE)
```

