---
title: "First Year Project: Extinction prediciton"
output: pdf_document
date: "2023-02-07"
---

```{r setup, include=FALSE, results=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggpmisc")) install.packages("ggpmisc")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("ggsignif")) install.packages("ggsignif")
library("ggplot2")
library("ggpmisc")
library("ggpubr")
library("ggsignif")
library(tibble)
library(ggplot2)
library(broom)
```

We start by importing the dataset directly from the csv file and saving it to the data variable:

```{r}
head(data <- read.csv('Factors Affecting Extinction.csv', header=T))
```

# (1) Initial Plotting

There are four different combinations of Size and Status, LR, LM, SR and SM. If we want to find the correlation between *extinction time* as a function of *pairs*, we can make a regression line with *pairs* as the predictor value and *extinction time* as the predicted value.

```{r}
ggplot(data, aes(x = Pairs, y = Time)) +
  geom_point() +
  facet_grid(Size ~ Status) +
  theme(legend.position = "top") +
  geom_smooth(method = "lm", formula = y ~ x) +
  
  stat_poly_eq(formula = y ~ x,
  aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
  parse = TRUE) +
  
  labs(title = "Raw data") +
  theme_bw()
```

# (2) Residual plot of raw (not transformed) data

```{r}
ggplot(data,
  mapping = aes(x = Pairs,
                y = resid(lm(Time ~ Pairs, data = data)))) +
  geom_point() +
  facet_grid(Size ~ Status) +
  geom_hline(yintercept = 0, color = "red") +
  xlab("Pairs") +
  ylab("Residuals") +
  labs(title = "Residual Plot (raw data)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom")
```

# (3) Transformations of time to log_time, sqrt_time, inverse_time
We now try different transformations on the data to see if we can get a better fitted regression line. We try 3 different transformation: log("time"), sqrt("time") and 1/("time").  

```{r}
data$log_time <- log2(data$Time)
data$sqrt_time <- sqrt(data$Time)
data$inverse_time <- 1/data$Time
head(data)
```

Plotting of log_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data, aes(x = Pairs, y = log_time)) +
  geom_point() +
  facet_grid(Size ~ Status) +
  theme(legend.position = "top") +
  geom_smooth(method = "lm", formula = y ~ x) +
  
  stat_poly_eq(formula = y ~ x,
  aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
  parse = TRUE) +
  
  labs(title = "Residual Plot (log_time)") +
  theme_bw()
```

Residual plot of log_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data,
  mapping = aes(x = Pairs,
                y = resid(lm(log_time ~ Pairs, data = data)))) +
  geom_point() +
  facet_grid(Size ~ Status) +
  geom_hline(yintercept = 0, color = "red") +
  xlab("Pairs") +
  ylab("Residuals") +
  labs(title = "Residual Plot (log_time)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom")
```

Plotting of sqrt_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data, aes(x = Pairs, y = sqrt_time)) +
  geom_point() +
  facet_grid(Size ~ Status) +
  theme(legend.position = "top") +
  geom_smooth(method = "lm", formula = y ~ x) +
  
  stat_poly_eq(formula = y ~ x,
  aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
  parse = TRUE) +
  
  labs(title = "Sqrt_time") +
  theme_bw()
```

Residual plot of sqrt_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data,
  mapping = aes(x = Pairs,
                y = resid(lm(sqrt_time ~ Pairs, data = data)))) +
  geom_point() +
  facet_grid(Size ~ Status) +
  geom_hline(yintercept = 0, color = "red") +
  xlab("Pairs") +
  ylab("Residuals") +
  labs(title = "Residual Plot (sqrt_time)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom")
```

Plotting of inverse_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data, aes(x = Pairs, y = inverse_time)) +
  geom_point() +
  facet_grid(Size ~ Status) +
  theme(legend.position = "top") +
  geom_smooth(method = "lm", formula = y ~ x) +
  
  stat_poly_eq(formula = y ~ x,
  aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
  parse = TRUE) +
  
  labs(title = "Inverse_time") +
  theme_bw()
```
Residual plot of inverse_time:

```{r echo=FALSE, results=FALSE, message=FALSE}
ggplot(data,
  mapping = aes(x = Pairs,
                y = resid(lm(inverse_time ~ Pairs, data = data)))) +
  geom_point() +
  facet_grid(Size ~ Status) +
  geom_hline(yintercept = 0, color = "red") +
  xlab("Pairs") +
  ylab("Residuals") +
  labs(title = "Residual Plot (inverse_time)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom")
```
As we can see, the downward trend of the points is no longer apparent. 
